An empirical assessment of strategies to model opponent effects in crash severity analysis

### 2020-02-17 Interactions-Crash-Severity-v1/Interactions-Crash-Severity-v1.tex
First version for submission, after comments from HH and MF

  \begin{abstract}
  Road accidents impose an important burden on health and the economy.
  Numerous efforts to understand the factors that affect road collisions
  have been undertaken. One stream of research focus on modelling the
  severity of crashes. Crash severity research is useful to clarify the
  way different factors can influence the outcome of an event. The
  objective of this paper is to assess different strategies to model the
  interactions between participants in a crash in the context of crashes
  involving two parties. Towards this objective, a series of models are
  estimated using data from Canada's National Collision Database. Three
  levels of crash severity (no injury/injury/fatality) are analyzed using
  ordered logit models and covariates for the participants in the crash
  and the conditions of the crash. Modelling strategies include different
  ways of introducing the covariates (e.g., in a single-level or
  multi-level form), as well as by subsetting the dataset. The models are
  assessed using predicted shares and classes of outcomes, and the results
  highlight the importance of considering opponent effects in crash
  severity analysis. The study also suggests that hierarchical (i.e.,
  multi-level) specifications and subsetting do not necessarily perform
  better than a relatively simple single-level model with opponent
  effects. The results of this study provide insights regarding the
  performace of different modelling strategies, and should be informative
  to researchers working with crash severity models.
  \end{abstract}

@@

Road safety continues to be a world-wide concern. According to a recent
report from the World Health Organization (2019), road accidents are the
8th leading cause of death for all ages, and the number one cause of
death for children and young people between the ages of 5 to 29. Of all
leading causes of death, road accidents are the only cause of death
unrelated to disease, health disorder, or infection. For this reason,
road accidents impose a heavy burden on individuals and society as a
whole. Gobally, the rate of road collision-related deaths per 100,000
population and 100,000 vehicles have both fallen, even as the number of
vehicles has grown (World Health Organization, 2019, Figs. 1 and 2).
These gains, although they are to be celebrated, cannot distract from
the crushing economic cost of premature death (e.g., Symons et al.,
2019; Wijnen et al., 2019), not to mention the long-term consequences
for survivors, measured in sometimes crippling emotional and physical
pain (e.g., Merlin et al., 2007; Devlin et al., 2019; Pelissier et al.,
n.d.).

Evidence from across the world suggests that the burden of road
accidents is not borne evenly. There are important disparities at the
international level, where the odds of death due to road crashes are
three times higher in low-income countries compared to high-income
countries; in fact, no reductions in road accident-related fatalities
were appreciated in low-income countries between 2013 and 2016 (World
Health Organization, 2019). In the case of high-income countries, where
substantial gains in road safety have been observed for years, said
gains have also been unevenly distributed; thus, while fatal crashes
involving older adults in the United States and Great Britain declined
between 1997 and 2010 (despite the graying of the population), the trend
remained stable or increased slightly in Australia in roughly the same
period (Thompson et al., 2018). There are also systematic differences in
the impact of road accidents. For example, in a study in the United
States, Obeng (2011) reported that the impact of covariates of crash
severity varied substantially in magnitude by gender. More recently,
Regev et al.~(2018) used adjusted crash risk to find that the risk of
crashes in Great Britain peaked for people 21 to 29 years of age; on the
other hand, the risk of fatal injuries for older drivers was constant,
irrespective of the seriousness of the crash - which highlights the
perils of accidents at older ages. Other studies have concentrated on
the consequences of road accidents for the young (e.g., Peek-Asa et al.,
2010), the old (e.g., Rakotonirainy et al., 2012), as well as
pedestrians and cyclists (e.g., Hanson et al., 2013; McArthur et al.,
2014).

Given the relevance and cost of this matter, as well as the important
variations of the impacts among different population segments, numerous
efforts have been conducted to better understand the factors that affect
road safety - including the probable consequences of crashes. Along
these lines, a stream of research in the analysis of road accidents is
concerned with the severity of crashes. In particular, multivariate
analysis of crash severity is a useful way to clarify the way various
factors can affect the outcome of an incident, and to discriminate
between various levels of injury, from no injury (i.e., property damage
only), to different degrees of injury up to and including death. This is
an active area of research, and one where methodological developments
have aimed at improving the reliability, accuracy, and precision of
models (e.g., Savolainen et al., 2011; Bogue et al., 2017; Shamsunnahar
and Eluru, 2013).

This paper aspires to contribute to the literature on crash severity by
assessing different modelling strategies useful to incorporate opponent
effects in crash analysis, in the context of incidents involving two
parties. The importance of these interactions has been recognized in the
existing literature (e.g., Chiou et al., 2013; Lee and Li, 2014; Li et
al., 2017; Tarrao et al., 2014), and a number of different modelling
strategies have been proposed. In this paper we present a systematic
assessment of several relevant modelling strategies, ranging from the
way variables are defined in single-level models, in multi-level models
(i.e., hierarchical models), as well as using data subsetting
approaches. For the assessment we use data from Canada's National
Collision Database, a database that collects all police-reported
collisions in the country. Using the most recent version of the dataset
(2017), three levels of crash severity (no injury/injury/fatality) are
analyzed using ordered logit models and covariates for the participants
in the crash and the conditions of the crash. For model assessment, we
conduct an in-sample prediction exercise using the estimation sample
(i.e., \emph{nowcasting}), and also an out-of-sample prediction exercise
using the dataset corresponding to 2016 (i.e., \emph{backcasting}). The
models are assessed using predicted shares and predicted classes of
outcomes at the individual level, using an extensive array of
verification statistics. The results highlight the importance of
considering opponent effects in crash severity analysis to improve the
goodness-of-fit and predictive performance o. On the other hand, the
study suggests that hierarchical variable specifications and subsetting
do not perform necessarily better than a relatively simple single-level
model with opponent effects.

The rest of this paper is structured as follows. In Section
\ref{sec:review-of-methods} we present a concise review of the methods
used to analyze crash severity, with a particular focus on techniques
that consider the effect of opponents in a crash. Section
\ref{sec:application} describes the data requirements, data
preprocessing, and the modelling strategies, along with the results of
model estimation. The results of assessing the models and the discussion
of these results is found in Section \ref{sec:assessment}. We then
present some additional thoughts about the applicability of this
approach in Section \ref{sec:further-considerations} before offering
some concluding remarks in Section \ref{sec:concluding-remarks}.

\hypertarget{sec:review-of-methods}{%
\section{Methodological approaches in crash severity
analysis}\label{sec:review-of-methods}}

\hypertarget{sec:general-considerations}{%
\subsection{General considerations}\label{sec:general-considerations}}

Modelling the outcomes of crashes in terms of the severity of injuries
to participants has been a preoccupation of transportation researchers,
planners, auto insurance companies, governments, and the general public
for decades. One of the earliests studies to investigate the severity of
injuries conditional on an accident having occurred was by White and
Clayton (1972). Kim et al.~(1995) later stated that the ``linkages
between severity of injury and driver characteristics and behaviors have
not been thoroughly investigated'' (p.~470). Nowadays, there is a
burgeoning literature on this subject, including methodological
developments, case studies, and more niche research with a focus on
particular situations (e.g., crashes at intersections, Mussone et al.,
2017; crashes in rural roads, Gong and Fan, 2017), and crashes involving
special population groups (e.g., crashes involving motorcyclists or
active travelers; see Shaheed et al., 2013; Salon and McIntyre, 2018).

Crash severity is often modelled using models for discrete outcomes. An
analyst interested in crash severity has at their disposal an ample menu
of models to choose from, including classification techniques from
machine learning (e.g., Iranitalab and Khattak, 2017; Chang and Wang,
2006; Effati et al., 2015; Khan et al., 2015), Poisson models for counts
(e.g., Ma et al., 2008), unordered logit/probit models (e.g., Tay et
al., 2011), as well as ordered logit/probit models (e.g., Rifaat and
Chin, 2007), with numerous variants, such as random parameters/mixed
logit (e.g., Aziz et al., 2013; Haleem and Gan, 2013), partial
proportional odds models (e.g., Mooradian et al., 2013; Sasidharan and
Menendez, 2014), and the use of copulas (e.g., Wang et al., 2015).
Recent reviews of methods include Savolainen et al.~(2011), Shamsunnahar
and Eluru (2013), and Mannering et al.~(2016).

Irrespective of the modelling framework employed, models of crash
severity often include variables in several categories, as shown with
examples in Table \ref{tab:variable-categories} (also see Montella et
al., 2013). Many crash databases and analyses also account for the
multievent nature of many crashes. Participants may have had different
roles in a crash depending on their context, with some acting as
operators of a vehicle (i.e., drivers, bicyclists), while others were
passengers. They also may differ depending on what type of traffic unit
they were, for example occupants of a light duty vehicle or a truck,
motorcyclists, or pedestrians. The multiplicity of roles makes for
complicated modelling decisions when trying to understand the severity
of injuries; for example, what is the unit of analysis, the person, the
traffic unit, or the collision? Not surprisingly, it is possible to find
examples of studies that adopt different perspectives. A common
simplifying strategy in model specification is to consider only
\emph{drivers} and/or only \emph{single-vehicle} crashes (e.g., Kim et
al., 2013; Gong and Fan, 2017; Lee and Li, 2014; Osman et al., 2018).
This strategy reduces the dimensions of the event, and it becomes
possible, for example, to equate the traffic unit to the person for
modelling purposes.

The situation becomes more complex when dealing with events that involve
two traffic units (e.g., Tarrao et al., 2014; Wang et al., 2015) and
multi-traffic unit crashes (e.g., Wu et al., 2014; Bogue et al., 2017).
Different strategies have been developed to study these, more complex
cases. A number of studies advocate the estimation of separate models
for different participants and/or situations. In this way, Wang and
Kockelman (2005) estimated models for single-vehicle and two-vehicle
crashes, while Savolainen and Mannering (2007) estimated models for
single-vehicle and multi-vehicle crashes. More recently, Duddu et
al.~(2018) and Penmetsa et al.~(2017) presented research that estimated
separate models for at-fault and not-at-fault drivers. The strategy of
estimating separate models also relies on subsetting the dataset,
although it is possible to link the relevant models more tightly by
means of a common covariance structure, as is the case of bivariate
models (e.g., Chiou et al., 2013; Chen et al., 2019) or models with
copulas (e.g., Rana et al., 2010; Shamsunnahar et al., 2014; Wang et
al., 2015).

A related strategy to specify a crash severity model is to organize the
data in such a way that it is possible to model the influence of the
attributes of the opponent in a crash. There are numerous examples of
studies that consider at least some characteristics of the oponents in
two- or multi-vehicle crashes. For example, Wang and Kockelman (2005)
considered the type of the opposing vehicle in their model for
two-vehicle collisions. Similarly, Tarrao et al.~(2014) included in
their analysis the age, wheelbase, weight, and engine size of the
opposing vehicle, while Bogue et al.~(2017) used the body type of the
opposing vehicle. Penmetsa et al.~(2017) and Duddu et al.~(2018) are two
of the most comprehensive examples of using opponent's information, as
they used individual attributes of opponents (their physical condition,
sex, and age), as well as characteristics of the opposing traffic unit
(the vehicle type of the opponent). The twin strategies of subsetting
the sample and using the attributes of the opponent are not mutually
exclusive, but neither are they used consistently together, as a scan of
the literature reveals.

@@

\caption{\label{tab:table-variable-categories}\label{tab:variable-categories}Categories of variables used in the analysis of crash severity with examples}
\centering
\fontsize{7}{9}\selectfont
\begin{tabular}[t]{l>{\raggedright\arraybackslash}p{22em}}
\toprule
Category & Examples\\
\midrule
Person-related & Attributes of participants in the crash, e.g., injury status, age, gender, licensing status, professional driver status\\
Traffic unit-related & Attributes of the traffic unit, e.g., type of traffic unit (car, motorcycle, etc.), maneouver, etc.\\
Crash-related & Attributes of the crash, e.g., location, weather conditions, light conditions, number of parties, etc.\\
Road-related & Attributes of the road, e.g., surface condition, grade, geometry, etc.\\
Opponent-related & Attributes of the opponent, e.g., age of opponent, gender of opponent, opponent vehicle type, etc.\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{sec:techniques}{%
\subsection{Modelling techniques}\label{sec:techniques}}

With respect to model structures, Shamsunnahar and Eluru (2013)
conducted an extensive comparison of models for discrete outcomes and
found only small differences in the performance of unordered models and
ordered models; however, ordered models are usually more parsimonious
since only one latent functions needs to be estimated for all outcomes,
as opposed to one for each outcome in unordered modelling mechanisms.
Bogue et al.~(2017) also compared unordered and ordered models in the
form of the mixed multinomial logit and a modified rank ordered logit,
respectively, and found that the ordered model performed best. To keep
the empirical assessment managable we will consider only the ordinal
logit model, and will comment on potential extensions in Section
\ref{sec:further-considerations}.

@@

The left-hand side of the expression above (\(y_{itk}^*\)) is a latent
(unobservable) variable that is associated with the severity of crash
\(k\) (\(k=1,\cdots,K\)) for participant \(i\) in traffic unit \(t\).
The right-hand side of the expression is split in four parts. The first
part collects \(l=1,\cdots,L\) individual attributes \(p\) for
participant \(i\) in traffic unit \(t\) and crash \(k\); these could
relate to the person (e.g., age, gender, and road user class). The
second part collects \(m=1,\cdots,M\) attributes \(u\) related to
traffic unit \(t\) in crash \(k\); these could be items such as maneuver
or vehicle type. The third part collects \(q=1,\cdots,Q\) attributes
\(c\) related to the crash \(k\), including crash-related and
road-related data, such as weather conditions, road alignment, and type
of surface. Lastly, the fourth element is a random term specific to
participant \(i\) in traffic unit \(t\) and crash \(k\). The function
consists of a total of \(Z=L+M+Q\) covariates and associated parameters.

@@

The latent variable is not observed directly, but it is possible to
posit a probabilistic relationship with the outcome \(y_{itk}\) (the
severity of crash \(k\) for participant \(i\) in traffic unit \(t\)).
Depending on the characteristics of the data and the assumptions made
about the random component of the latent function different models can
be obtained. For example, if crash severity is coded as a binary
variable (e.g., non-fatal/fatal), we can relate the latent variable to
the outcome as follows:

\begin{equation}
\label{eq:latent-function-2-outcomes}
y_{itk} = 
\begin{cases}
\text{fatal} & \text{if } y_{itk}^*>0\\
\text{non-fatal} & \text{if } y_{itk}^*\leq0
\end{cases}
\end{equation}

Due to the stochastic nature of the latent function, the outcome of the
crash is not fully determined. However, we can make the following
probability statement:

\begin{equation}
\label{eq:probability-2-outcomes}
P(y_{itk} = \text{fatal}) = P(y_{itk}^* > 0)
\end{equation}

In other words, the probability that individual \(i\) in traffic unit
\(t\) and crash \(k\) was a fatality equals the probability that the
latent variable is greater than zero. This implies (see Maddala, 1986,
p. 22):

\begin{equation}
\label{eq:probability-2-outcomes-2}
\begin{array}{rl}\
P(y_{itk} = \text{fatal}) &= P(\sum_{z=1}\theta_zp_{itkz} + \epsilon_{itk} > 0)\\ 
&=P(\epsilon_{itk} > -\sum_{z=1}\theta_zp_{itkz})
\end{array}
\end{equation}

If the random terms \(\epsilon_{itk}\) are assumed to follow the
logistic distribution, then the binary logit model is obtained; if they
are assumed to follow the normal distribution, the binary probit model
is obtained. More often, though, the outcome is recorded using more
categories, for example property damage only (PDO)/injury/fatality. A
similar approach can be adopted, with a latent variable that relates to
the outcome as follows:

\noindent where \(\mu_1\) and \(\mu_2\) are estimable thresholds. In
this case, the associated probability statements are as follows:

\hypertarget{sec:strategies}{%
\subsection{Model specification strategies}\label{sec:strategies}}

In this paper we consider three general model specification strategies,
as follows:

\begin{itemize}
\tightlist
\item
  Strategy 1. Introducing opponent-related variables
\item
  Strategy 2. Single-level model and multi-level (hierarchical) model
  specifications
\item
  Strategy 3. Full sample and sample subsetting
\end{itemize}

Introduction of opponent related-variables was explained in the
preceding subsection. In this way, a base model that ignores opponent
effects is given by Equation \ref{eq:latent-function}. Strategy 1, in
contrast, is Equation \ref{eq:latent-function-with-opponent-variables},
which includes opponent-related variables. These two equations are also
examples of single-level models. Next we describe Strategies 2 and 3.

\hypertarget{strategy-2-hierarchical-model-specification}{%
\subsubsection{Strategy 2: hierarchical model
specification}\label{strategy-2-hierarchical-model-specification}}

@@

Alterted readers will identify this model specification strategy as
Casetti's expansion method (Casetti, 1972; Roorda et al., 2010). This is
a deterministic strategy for modelling contextual effects which, when
augmented with random components, becomes the well-known multi-level
modelling method (Hedeker and Gibbons, 1994, more on this in Section
\ref{sec:further-considerations}). It is worthwhile to note that
higher-order hierarchical effects are possible; for instance, individual
attributes nested within traffic units, which in turn are nested within
collisions. We do not explore higher-level hierarchies further in the
current paper.

\hypertarget{strategy-3-sample-subsetting}{%
\subsubsection{Strategy 3: sample
subsetting}\label{strategy-3-sample-subsetting}}

@@

In this section we present the setting for the empirical assessment of
the modelling strategies discussed in Section
\ref{sec:review-of-methods}, namely matters related to data and model
estimation.

@@

The NCDB is available from 1999. For the purpose of this paper, we use
the data corresponding to 2017, which is the most recent year available
as of this writing. Furthermore, for assessment we also use the data
corresponding to 2016. Similar to databases in other jurisdictions (see
Montella et al., 2013), the NCDB contains information pertaining to the
collision, the traffic unit(s), and the person(s) involved in a crash,
as shown in Tables \ref{tab:ncdb-descriptives-collision},
\ref{tab:ncdb-descriptives-vehicle}, and
\ref{tab:ncdb-descriptives-person}. Notice that, compared to Table
\ref{tab:variable-categories}, crash-related variables and road-related
variables are collected under a single variable class, namely
collision-related, since they are unique for each crash.

@@

\multicolumn{3}{l}{Source NCDB available from https://open.canada.ca/data/en/dataset/1eb9eba7-71d1-4b30-9fb1-30cbdab7e63a}\\

\begin{longtable}[t]{ll>{\raggedright\arraybackslash}p{32em}}
\caption{\label{tab:ncdb-descriptives-vehicle}\label{tab:ncdb-descriptives-vehicle}Contents of National Collision Database: Traffic unit-level variables}\\
\toprule
Variable & Description & Notes\\
\midrule
\endfirsthead
\caption[]{\label{tab:ncdb-descriptives-vehicle}Contents of National Collision Database: Traffic unit-level variables \textit{(continued)}}\\
\toprule
Variable & Description & Notes\\
\midrule
\endhead
\
\endfoot
\bottomrule
\multicolumn{3}{l}{\textit{Note: }}\\
\multicolumn{3}{l}{Source NCDB available from https://open.canada.ca/data/en/dataset/1eb9eba7-71d1-4b30-9fb1-30cbdab7e63a}\\
\multicolumn{3}{l}{Source data files for analysis also available from https://drive.google.com/open?id=12aJtVBaQ4Zj0xa7mtfqxh0E48hKCb\_XV}\\
\endlastfoot
\rowcolor{gray!6}  V\_ID & Vehicle sequence number & Number of vehicles: 1-98; Pedestrian sequence number: 99; unknown.\\
V\_TYPE & Vehicle type & 21 levels: Light Duty Vehicle (Passenger car, Passenger van, Light utility vehicles and light duty pick up trucks); Panel/cargo van (<= 4536 KG GVWR   Panel or window type of van designed primarily for carrying goods); Other trucks and vans (<= 4536 KG GVWR); Unit trucks (> 4536 KG GVWR); Road tractor; School bus; Smaller school bus (< 25 passengers); Urban and Intercity Bus; Motorcycle and moped; Off road vehicles; Bicycle; Purpose-built motorhome; Farm equipment; Construction equipment; Fire engine; Snowmobile; Street car; Data element is not applicable  (e.g. dummy vehicle record created for pedestrian); Choice is other than the preceding values; unknown; not reported by jurisdiction.\\
\rowcolor{gray!6}  V\_YEAR & Vehicle model year & Model year; dummy for pedestrians; unknown; not reported by jurisdiction.\\*
\end{longtable}
\endgroup{}

>>>>

  \begin{abstract}
  Road accidents impose an important burden on health and the economy.
  Numerous efforts have been undertaken to understand the factors that
  affect road collisions in general, and the severity of crashes in
  particular. In this literature several strategies have been proposed to
  model opponent effects, that is, ways to deal with the way the different
  parties in a crash interact to influence the severity of the individual
  outcomes. These strategies include the use of opponent variables, data
  subsetting, and estimating models with hierarchical components. Since,
  no systematic assessment has been conducted of the performance of these
  strategies, they appear to be used in an ad-hoc fashion in the
  literature. The objective of this paper is to empirically evaluate ways
  to model opponent effects in the context of crashes involving two
  parties. To this end, a series of models are estimated using data from
  Canada's National Collision Database. Three levels of crash severity (no
  injury/injury/fatality) are analyzed using ordered logit models and
  covariates for the participants in the crash and the conditions of the
  crash. The models are assessed using predicted shares and classes of
  outcomes, and the results highlight the importance of considering
  opponent effects in crash severity analysis. The study also suggests
  that hierarchical (i.e., multi-level) specifications and subsetting do
  not necessarily perform better than a relatively simple single-level
  model with opponent-related factors. The results of this study provide
  insights regarding the performance of different modelling strategies,
  and should be informative to researchers in the field of crash severity.
  \end{abstract}

@@

Evidence from across the world suggests that the burden of road
accidents is not borne evenly. There are important disparities at the
international level, where the odds of death due to road crashes are
three times higher in low-income countries compared to high-income
countries (World Health Organization, 2019). In the case of high-income
countries, where substantial gains in road safety have been observed for
years, said gains have also been unevenly distributed; thus, while fatal
crashes involving older adults in the United States and Great Britain
declined between 1997 and 2010 (despite the graying of the population),
the trend remained stable or increased slightly in Australia in roughly
the same period (Thompson et al., 2018). The impact of road accidents
also varies considerably. For example, in a study in the United States,
Obeng (2011) reported that the factors affecting crash severity varied
substantially in magnitude by gender. More recently, Regev et al.~(2018)
used adjusted crash risk to find that the risk of fatal injuries for
older drivers remained high even for minor collisions - which highlights
the perils of accidents at older ages. Other studies have concentrated
on the consequences of road accidents for the young (e.g., Peek-Asa et
al., 2010), the old (e.g., Rakotonirainy et al., 2012), as well as
pedestrians and cyclists (e.g., Hanson et al., 2013; McArthur et al.,
2014).

Given the relevance and cost of this matter, as well as the unevenness
of impacts among different population segments, numerous efforts have
been conducted to better understand the factors that affect road safety
- including the probable consequences of crashes. Along these lines, a
stream of research in the analysis of road accidents is concerned with
the severity of crashes. In particular, multivariate analysis of crash
severity is a useful way to clarify the way various factors can affect
the outcome of an incident, and to discriminate between various levels
of injury. This is an active area of research, and one where
methodological developments have aimed at improving the reliability,
accuracy, and precision of models (e.g., Savolainen et al., 2011; Bogue
et al., 2017; Shamsunnahar and Eluru, 2013). Of interest in the
literature on crash severity is the way different parties in a crash
interact to influence the severity of individual outcomes. The
importance of these interactions has been recognized in the existing
literature (e.g., Chiou et al., 2013; Lee and Li, 2014; Li et al., 2017;
Torrao et al., 2014), and a number of different modelling strategies
have been proposed. These include the use of \emph{data subsetting},
\emph{opponent variables}, and estimating models with \emph{hierarchical
components}. A gap in the literature, however, is a systematic
assessment of the performance of these various strategies.

The objective of this paper is to empirically evaluate different
strategies to model opponent effects in crash severity, in the context
of incidents involving two parties. Therefore, we present a systematic
assessment of several relevant modelling strategies. For the assessment
we use data from Canada's National Collision Database, a database that
collects all police-reported collisions in the country. Using the most
recent version of the data set (2017). Three levels of crash severity
(no injury/injury/fatality) are analyzed using ordered logit models, and
covariates for the participants in the crash and the conditions of the
crash. For model assessment, we conduct an in-sample prediction exercise
using the estimation sample (i.e., \emph{nowcasting}), and also an
out-of-sample prediction exercise using the data set corresponding to
2016 (i.e., \emph{backcasting}). The models are assessed using predicted
shares and predicted classes of outcomes at the individual level, using
an extensive array of verification statistics.

The rest of this paper is structured as follows. In Section
\ref{sec:background} we discuss some background matters, and follow this
with a concise review of the modelling strategies used to analyze crash
severity \ref{sec:methods}. Section \ref{sec:application} describes the
data requirements, data preprocessing, and the modelling strategies,
along with the results of model estimation. The results of assessing the
models and the discussion of these results is found in Section
\ref{sec:assessment}. We then present some additional thoughts about the
applicability of this approach in Section
\ref{sec:further-considerations} before offering some concluding remarks
in Section \ref{sec:concluding-remarks}.

\hypertarget{sec:background}{%
\section{Background}\label{sec:background}}

Modelling the severity of injuries to participants in road crashes has
been a preoccupation of transportation researchers, planners, auto
insurance companies, governments, and the general public for decades.
One of the earliest studies to investigate the severity of injuries
conditional on an accident having occurred was by White and Clayton
(1972). Kim et al.~(1995) later stated that the ``linkages between
severity of injury and driver characteristics and behaviors have not
been thoroughly investigated'' (p.~470). Nowadays, there is a burgeoning
literature on this subject, including methodological developments, case
studies, and more niche research with a focus on particular situations
(e.g., crashes at intersections, Mussone et al., 2017; crashes in rural
roads, Gong and Fan, 2017), and crashes involving special population
groups (e.g., crashes involving motorcyclists or active travelers; see
Shaheed et al., 2013; Salon and McIntyre, 2018).

Crash severity is often modeled using models for discrete outcomes. An
analyst interested in crash severity has at their disposal an ample
repertoire of models to choose from, including classification techniques
from machine learning (e.g., Iranitalab and Khattak, 2017; Chang and
Wang, 2006; Effati et al., 2015; Khan et al., 2015), Poisson models for
counts (e.g., Ma et al., 2008), unordered logit/probit models (e.g., Tay
et al., 2011), as well as ordered logit/probit models (e.g., Rifaat and
Chin, 2007), with numerous variants, such as random parameters/mixed
logit (e.g., Aziz et al., 2013; Haleem and Gan, 2013), partial
proportional odds models (e.g., Mooradian et al., 2013; Sasidharan and
Menendez, 2014), and the use of copulas (e.g., Wang et al., 2015).
Recent reviews of methods include Savolainen et al.~(2011), Shamsunnahar
and Eluru (2013), and Mannering et al.~(2016).

Irrespective of the modelling framework employed, models of crash
severity often include variables in several categories, as shown with
examples in Table \ref{tab:variable-categories} (also see Montella et
al., 2013). Many crash databases and analyses also account for the
multievent nature of many crashes. Participants may have had different
roles in a crash depending on their context, with some acting as
operators of a vehicle (i.e., drivers, bicyclists), while others were
passengers. They also may differ depending on what type of traffic unit
they were, for example pedestrians, or operators/passengers of a
vehicle. The multiplicity of roles makes for complicated modelling
decisions when trying to understand the severity of injuries; for
example, what is the unit of analysis, the person, the traffic unit, or
the collision? Not surprisingly, it is possible to find examples of
studies that adopt different perspectives. A common simplifying strategy
in model specification is to consider only \emph{drivers} and/or only
\emph{single-vehicle} crashes (e.g., Kim et al., 2013; Gong and Fan,
2017; Lee and Li, 2014; Osman et al., 2018). This strategy reduces the
dimensions of the event, and it becomes possible, for example, to equate
the traffic unit to the person for modelling purposes.

The situation becomes more complex when dealing with events that involve
two traffic units (e.g., Torrao et al., 2014; Wang et al., 2015) and
multi-traffic unit crashes (e.g., Wu et al., 2014; Bogue et al., 2017).
Different strategies have been developed to study these, more complex
events. A number of studies advocate the estimation of separate models
for different participants and/or situations. In this way, Wang and
Kockelman (2005) estimated models for single-vehicle and two-vehicle
crashes, while Savolainen and Mannering (2007) estimated models for
single-vehicle and multi-vehicle crashes. More recently, Duddu et
al.~(2018) and Penmetsa et al.~(2017) presented research that estimated
separate models for at-fault and not-at-fault drivers. The strategy of
estimating separate models relies on \emph{subsetting} the data set,
although it is possible to link the relevant models more tightly by
means of a common covariance structure, as is the case of bivariate
models (e.g., Chiou et al., 2013; Chen et al., 2019) or models with
copulas (e.g., Rana et al., 2010; Shamsunnahar et al., 2014; Wang et
al., 2015).

A related strategy to specify crash severity models is to organize the
data in such a way that it is possible to introduce \emph{opponent
effects}. There are numerous examples of studies that consider at least
some characteristics of the opponents in two- or multi-vehicle crashes.
For example, Wang and Kockelman (2005) considered the type of the
opposing vehicle in their model for two-vehicle collisions. Similarly,
Torrao et al.~(2014) included in their analysis the age, wheelbase,
weight, and engine size of the opposing vehicle, while Bogue et
al.~(2017) used the body type of the opposing vehicle. Penmetsa et
al.~(2017) and Duddu et al.~(2018) are two of the most comprehensive
examples of using opponent's information, as they used individual
attributes of opponents (their physical condition, sex, and age), as
well as characteristics of the opposing traffic unit (the vehicle type
of the opponent). The twin strategies of subsetting the sample and using
the attributes of the opponent are not mutually exclusive, but neither
are they used consistently together, as a scan of the literature
reveals. Another strategy is to introduce \emph{hierarchical components}
in the model, a technique widely used in the hierarchical or multi-level
modelling literature.

In this paper we consider three general modelling strategies, as
follows:

\begin{itemize}
\tightlist
\item
  Strategy 1. Introducing opponent-related factors
\item
  Strategy 2. Single-level model and multi-level (hierarchical) model
  specifications
\item
  Strategy 3. Full sample and sample subsetting
\end{itemize}

These are discussed in more detail in the following section.

@@

\caption{\label{tab:table-variable-categories}\label{tab:variable-categories}Categories of variables used in the analysis of crash severity with examples}
\centering
\fontsize{7}{9}\selectfont
\begin{tabular}[t]{l>{\raggedright\arraybackslash}p{22em}}
\toprule
Category & Examples\\
\midrule
Human-factors & Attributes of participants in the crash, e.g., injury status, age, gender, licensing status, professional driver status\\
Traffic unit-factors & Attributes of the traffic unit, e.g., type of traffic unit (pedestrian, car, motorcycle, etc.), maneuver, etc.\\
Environmental-factors & Attributes of the crash, e.g., location, weather conditions, light conditions, number of parties, etc.\\
Road-factors & Attributes of the road, e.g., surface condition, grade, geometry, etc.\\
Opponent-related factors & Attributes of the opponent, e.g., age of opponent, gender of opponent, opponent vehicle type, etc.\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{sec:methods}{%
\section{Methods}\label{sec:methods}}

\hypertarget{choice-of-model}{%
\subsection{Choice of model}\label{choice-of-model}}

Shamsunnahar and Eluru (2013) conducted an extensive comparison of
models for discrete outcomes in crash severity analysis, and found only
small differences in the performance of unordered models and ordered
models; however, ordered models are usually more parsimonious since only
one latent functions needs to be estimated for all outcomes, as opposed
to one for each outcome in unordered models. Bogue et al.~(2017) also
compared unordered and ordered models in the form of the mixed
multinomial logit and a modified rank ordered logit, respectively, and
found that the ordered model performed best. To keep the empirical
assessment manageable, in this paper we will consider only the ordinal
logit model, and will comment on potential extensions in Section
\ref{sec:further-considerations}.

@@

The left-hand side of the expression above (\(y_{itk}^*\)) is a latent
(unobservable) variable that is associated with the severity of crash
\(k\) (\(k=1,\cdots,K\)) for participant \(i\) in traffic unit \(t\).
The right-hand side of the expression is split in four parts. The first
part collects \(l=1,\cdots,L\) human-factors \(p\) for participant \(i\)
in traffic unit \(t\) and crash \(k\); these could relate to the person
(e.g., age, gender, and road user class). The second part collects
\(m=1,\cdots,M\) attributes \(u\) related to traffic unit \(t\) in crash
\(k\); these could be items such as maneuver or vehicle type. The third
part collects \(q=1,\cdots,Q\) attributes \(c\) related to the crash
\(k\), including environmental-factors and road-factors, such as weather
conditions, road alignment, and type of surface. Lastly, the fourth
element is a random term specific to participant \(i\) in traffic unit
\(t\) and crash \(k\). The function consists of a total of \(Z=L+M+Q\)
covariates and associated parameters.

The latent variable is not observed directly, but it is possible to
posit a probabilistic relationship with the outcome \(y_{itk}\) (the
severity of crash \(k\) for participant \(i\) in traffic unit \(t\)).
Depending on the characteristics of the data and the assumptions made
about the random component of the latent function different models can
be obtained. For example, if crash severity is coded as a variable with
three outcomes (e.g., property damage only/injury/fatal), we can relate
the latent variable to the outcome as follows:

\noindent where \(\mu_1\) and \(\mu_2\) are estimable thresholds. Due to
the stochastic nature of the latent function, the outcome of the crash
is not fully determined. However, we can make the following probability
statements:

\hypertarget{strategy-1-opponent-related-factors}{%
\subsection{Strategy 1: opponent-related
factors}\label{strategy-1-opponent-related-factors}}

@@

\hypertarget{strategy-2-hierarchical-model-specification}{%
\subsection{Strategy 2: hierarchical model
specification}\label{strategy-2-hierarchical-model-specification}}

@@

Discerning readers will identify this model specification strategy as
Casetti's expansion method (Casetti, 1972; Roorda et al., 2010). This is
a deterministic strategy for modelling contextual effects which, when
augmented with random components, becomes the well-known multi-level
modelling method (Hedeker and Gibbons, 1994, more on this in Section
\ref{sec:further-considerations}). It is worthwhile to note that
higher-order hierarchical effects are possible; for instance, individual
attributes nested within traffic units, which in turn are nested within
collisions. We do not explore higher-level hierarchies further in the
current paper.

\hypertarget{strategy-3-sample-subsetting}{%
\subsection{Strategy 3: sample
subsetting}\label{strategy-3-sample-subsetting}}

@@

In this section we present the setting for the empirical assessment of
the modelling strategies discussed in Section \ref{sec:methods}, namely
matters related to data and model estimation.

@@

The NCDB is available from 1999. For the purpose of this paper, we use
the data corresponding to 2017, which is the most recent year available
as of this writing. Furthermore, for assessment we also use the data
corresponding to 2016. Similar to databases in other jurisdictions (see
Montella et al., 2013), the NCDB contains information pertaining to the
collision, the traffic unit(s), and the person(s) involved in a crash.
The definitions of variables in this database are shown in the Appendix
at the end of this document, in Tables
\ref{tab:ncdb-descriptives-collision},
\ref{tab:ncdb-descriptives-vehicle}, and
\ref{tab:ncdb-descriptives-person}. Notice that, compared to Table
\ref{tab:variable-categories}, environmental-factors variables and
road-factors variables are collected under a single variable class,
namely collision-related, since they are unique for each crash.

@@
